---
title: "Adapters and building your own"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adapters and building your own}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scans)
```

An adapter snapshots a completed run from one framework into the
`TrajectoryBundle` contract. All adapters share three rules:

- **Post-run and read-only.** An adapter reads a settled value. It never calls a
  provider, replays a tool, or reaches into live objects.
- **No invention.** When the source lacks something the contract can express,
  the adapter records a **loss** naming the field and the reason
  (`unsupported`, `redacted`, `truncated`, or `externalized`) rather than
  filling the gap.
- **Sanitized.** Sensitive named fields in tool arguments, results, and metadata
  are redacted before they enter a bundle, and the redaction is itself a loss.

`as_trajectory()` dispatches on the input's class. Sources that arrive as
unclassed lists or tibbles use an explicit `as_trajectory_*()` function so
scans never guesses what a bare list is.

## ellmer

`as_trajectory_ellmer()` converts a completed `Chat` or a list of public turn
objects. It is also the foundation the other adapters reuse for
model-visible turns.

```{r, eval = requireNamespace("ellmer", quietly = TRUE)}
request <- ellmer::ContentToolRequest("call-1", "lookup", list(id = 7L))
turns <- list(
  ellmer::UserTurn("Look up record 7"),
  ellmer::AssistantTurn(list(request)),
  ellmer::UserTurn(list(ellmer::ContentToolResult("found", request = request))),
  ellmer::AssistantTurn("Record 7 was found.")
)
bundle <- as_trajectory(turns)
trajectory_events(bundle)[, c("event_type", "name", "call_id", "status")]
```

Text, tool requests and results, thinking, and token usage become events.
Tool results link to their request through `call_id`; an ambiguous match is
left unlinked and reported by the `ambiguous_tool_correlation` scan rather
than guessed.

## OpenTelemetry GenAI spans

`as_trajectory_otel()` accepts the conversations returned by
`read_connect_traces()`, or a flat list of spans that it groups itself. The
most recent `chat` span of a conversation carries the whole message history;
`execute_tool` spans add each call's duration and failure. Non-GenAI span and
resource attributes are kept in trajectory metadata. See
`vignette("connect")`.

## vitals

`as_trajectory_vitals()` converts an evaluated vitals `Task` or the samples
returned by `vitals::vitals_log_read()`. Each sample and epoch becomes a
trajectory with a stable identifier, the solver chat is converted through the
ellmer adapter, and scores join as evaluations:

```{r, eval = FALSE}
bundle <- as_trajectory_vitals(task)
trajectory_evaluations(bundle)

samples <- vitals::vitals_log_read(log_path)
bundle <- as_trajectory_vitals(samples, source_uri = log_path)
```

## Deputy

A completed Deputy result adds run lifecycle events, tool timing, usage,
immutable run context, and delegation identity to its ellmer turns. Passing a
list of results keeps each run as its own trajectory, and a child is linked
through `parent_trajectory_id` only when its parent is in the same snapshot:

```{r, eval = FALSE}
bundle <- as_trajectory_deputy(list(parent_result, child_result))
trajectory_info(bundle)[, c("trajectory_id", "run_id", "parent_trajectory_id")]
```

Unresolved parent identifiers and missing run identifiers are recorded as
losses.

## dsprrr

A dsprrr module converts through its public trace exporter; an already
exported trace tibble uses the explicit function because it has no class of
its own:

```{r, eval = FALSE}
bundle <- as_trajectory(program)

traces <- dsprrr::export_traces(program, include_prompts = TRUE, include_outputs = TRUE)
bundle <- as_trajectory_dsprrr(traces)
```

Each trace becomes a trajectory with token, cost, latency, model, and output
context. The program artifact digest is the source identity; it establishes
integrity, not authenticity.

## commons

commons reconstructs conversations and their provenance from OpenTelemetry
traces. Read through commons, then convert the settled value:

```{r, eval = FALSE}
conversations <- commons::trajectory_read("path/to/otel-traces")
bundle <- as_trajectory_commons(conversations)
filter_trajectory_events(bundle, event_type = "commons:provenance")
```

Model-visible turns reuse the ellmer adapter. Activity time, the local or
Connect source descriptor, trust tags, and citation decisions stay
inspectable as `commons:provenance` events. Missing or conflicting provenance
becomes a loss; scans never infers a stronger trust claim.

## Tempest

Tempest exposes a bounded, deterministic review of a completed STORM or
Co-STORM product. The adapter uses that authoritative projection rather than
re-deriving it:

```{r, eval = FALSE}
review <- tempest::tempest_trajectory_review(result)
bundle <- as_trajectory_tempest(review)
filter_trajectory_events(bundle, event_type = "tempest:join")
```

The product is one trajectory. Stage order is authoritative; agent runs,
programs, knowledge, evidence identities, joins, and findings keep their
source order. Prompts, responses, source content, paths, and credentials are
excluded and recorded as losses.

## Build a bundle from your own framework

`TrajectoryBundle()` accepts rectangular data directly. Non-empty tables need
only a few columns; optional canonical columns are filled with typed missing
values and extra columns are kept.

| Table | Required columns |
|---|---|
| `trajectories` | `trajectory_id`, `source_type` |
| `turns` | `trajectory_id`, `turn_id`, `turn_index`, `role` |
| `events` | `trajectory_id`, `event_id`, `event_index`, `event_type` |
| `evaluations` | `trajectory_id`, `evaluation_id` |
| `losses` | `field`, `reason`, `detail` |

The scans rely on a handful of event columns: `event_type` values
`tool_call` and `tool_result`, the tool `name`, a `call_id` shared by a call and
its result, the canonical `value` list-column for tool-call arguments and
tool-result payloads, and `status` or `error` for failures.

```{r}
trajectories <- data.frame(
  trajectory_id = "run-1",
  source_type = "my-framework",
  status = "completed"
)

turns <- data.frame(
  trajectory_id = "run-1",
  turn_id = c("turn-1", "turn-2"),
  turn_index = 1:2,
  role = c("assistant", "user")
)

events <- tibble::tibble(
  trajectory_id = "run-1",
  turn_id = c("turn-1", "turn-2", "turn-2"),
  event_id = c("event-1", "event-2", "event-3"),
  event_index = 1:3,
  event_type = c("tool_call", "tool_result", "tool_result"),
  name = "search",
  call_id = c("call-1", "call-1", "call-2"),
  value = list(list(query = "Detroit"), "sunny", NULL),
  status = c("completed", "completed", "failed"),
  error = c(NA, NA, "timeout")
)

losses <- data.frame(
  trajectory_id = "run-1",
  field = "latency",
  reason = "unsupported",
  detail = "The framework does not record per-call latency."
)

bundle <- TrajectoryBundle(trajectories, turns, events, losses = losses)
scan_trajectories(bundle)[, c("severity", "label", "event_ids")]
trajectory_losses(bundle)
```

Validation errors name the table and column so a new adapter can be built up
incrementally. Once the tables are right, wrap the conversion in a function
and, if your framework's objects carry a class, register an `as_trajectory()`
method for it with `S7::method()`.
