---
title: "Reviewing Posit Connect traces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reviewing Posit Connect traces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Posit Connect can retain the OpenTelemetry traces that deployed applications
emit. ellmer follows the GenAI semantic conventions, so a Connect deployment
that uses ellmer leaves a complete record of every model call, its message
history, and each tool execution. scans reads that record directly and turns it
into trajectories you can scan and review, with no changes to the producing
application beyond enabling capture.

## Prerequisites on the producing application

1. **Server permission.** A Connect administrator enables OpenTelemetry and
   allows content instrumentation:

   ```ini
   [OpenTelemetry]
   Enabled = true
   AllowContentInstrumentation = true
   ```

2. **Content observability** is turned on for each application under
   **Settings > Monitoring**, and the application is restarted afterwards.

3. **Message content is captured.** ellmer records prompts and responses in its
   spans only when the process runs with
   `OTEL_INSTRUMENTATION_GENAI_CAPTURE_MESSAGE_CONTENT=true`. Set it as a
   Connect Var. Without it the reader finds spans but has no messages to
   reconstruct.

4. **Conversations are identified.** Spans are grouped by the
   `gen_ai.conversation.id` attribute, found by walking up each span's
   ancestry. commons agents set it; other ellmer apps should set the chat's
   conversation identifier or open a parent span carrying the attribute.
   Without it, each model call becomes its own one-turn trajectory.

Captured message content is sensitive. It stays on the Connect server, and the
review app described below reads it with the credentials of its deployment
owner.

## Read traces from R

`read_connect_traces()` needs `CONNECT_SERVER` and `CONNECT_API_KEY`, and the
key's user must own or collaborate on the content. It accepts a content GUID,
a content URL, or a dashboard URL.

```{r}
library(scans)

conversations <- read_connect_traces(
  "11111111-1111-4111-8111-111111111111",
  n = 200,
  from = Sys.Date() - 30
)
attr(conversations, "read_info")

bundle <- as_trajectory_otel(conversations)
summarize_trajectories(bundle)
scan_trajectories(bundle)
```

### Windows and limits

- `from` and `to` select conversations: a conversation is kept when one of
  its model calls started in `[from, to)`. A kept conversation then retains
  its earlier spans as far back as the read reached, which is an hour before
  `from`, so token totals, tool failures, and the start time describe the
  conversation, not just the window. Spans at or after `to` are dropped.
  Widen `from` if you need the timing and usage of calls made earlier than
  that hour. When both are omitted the reader takes the seven days ending
  now. Pass `NULL` explicitly to leave one side open. Bounds may be POSIXct,
  Date, or ISO 8601 strings; a string without a zone is read as UTC.
- `n` keeps the most recent conversations; `NULL` keeps all of them.
- `max_spans` caps how many GenAI spans are read. Only spans with `gen_ai.*`
  attributes count, so a Shiny application's reactive spans do not exhaust the
  budget before the conversations are reached.
- `jobs` controls whether retained per-job trace stores are read as well.
  Connect did not migrate traces recorded before its content-wide store, so
  the default reads both and de-duplicates. `jobs = FALSE` is much faster on
  deployments with many past jobs and is safe once the content-wide store
  holds everything you care about.

The `read_info` attribute records the window, how many spans and conversations
were found, whether the span ceiling or `n` left conversations out, and whether
a page failed to load. The review app shows the same information so a reviewer
can tell a quiet week from a truncated or incomplete read.

### How a conversation is reconstructed

Provider instrumentation records the whole message history on every model
call, so the most recent `chat` span of a conversation carries the complete
exchange. `execute_tool` spans contribute each tool call's duration and
failure, which the message history alone does not hold. Non-GenAI span and
resource attributes such as `enduser.id` or a session identifier are kept in
trajectory metadata and surface in the app as badges and search terms.

## Deploy the review app

`scans_app_connect()` builds a review app whose application switcher reads
from Connect. The repository's `deploy/connect/` directory contains a ready
`app.R` configured through Connect Vars, or call the function from your own:

```{r}
library(scans)

scans_app_connect(
  c(
    "Support assistant" = "11111111-1111-4111-8111-111111111111",
    "Research assistant" = "22222222-2222-4222-8222-222222222222"
  ),
  n = 100,
  jobs = FALSE
)
```

Only content GUIDs are accepted, so every source stays on the hosting
`CONNECT_SERVER` and a content URL cannot redirect the injected API key
elsewhere. Each application is read when first selected in any session. The
snapshot is then shared by every session of the deployment until it is thirty
minutes old or a reviewer presses **Reload traces**.

The app depends on shiny, bslib, htmltools, httr2, and jsonlite. Attach them in
`app.R` (as `deploy/connect/app.R` does) so Posit Publisher or rsconnect
bundles them.

### Annotations on Connect

Pass a `scans_annotations()` store to record reviewer labels and notes. Point
it at a file on storage that survives redeployment and that every process of
the app can write to, and run the app as a single process so all reviewers
append to one log:

```{r}
store <- scans_annotations("/data/scans-reviews/annotations.jsonl")
scans_app_connect(sources, annotations = store)
```

The bundled `app.R` reads `SCANS_ANNOTATIONS_DIR` for this. Without a store
the app makes no writes and shows no annotation controls.

## Security model

- **The app reads as its owner.** Connect injects an ephemeral API key for the
  deployment owner. Every viewer of the app therefore sees every configured
  application's conversations, including prompts, responses, tool arguments,
  and results. Restrict the app's viewers accordingly.
- **Failures are not echoed.** A source that fails to load is reported in the
  app as a failure with a fixed message; the underlying error, which may
  include hostnames or paths, goes to the server log.
- **Transcripts are sanitized.** Model-authored HTML is rebuilt against an
  element and attribute allowlist. Markup that cannot be shown safely appears
  as escaped source so the reviewer can still see what the model emitted.
- **Secrets are redacted before they reach a bundle.** Adapters redact
  sensitive named fields in tool arguments, results, and metadata. The
  transcript bounds every payload so an oversized result cannot stall the
  browser.
- **Nothing is written without a store.** The app never modifies a bundle. With
  an annotation store it appends records and nothing else.

## Agents built with commons

commons agents log their conversations with provenance tags and citation
decisions that the GenAI spans do not carry. To keep those, read through
commons and convert with `as_trajectory_commons()`, or pass
`reader = "commons"` to `scans_app_connect()`:

```{r}
conversations <- commons::trajectory_read("path-or-guid")
bundle <- as_trajectory_commons(conversations)

scans_app_connect(sources, reader = "commons")
```
