---
title: "Investigate a support assistant offline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigate a support assistant offline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

An order lookup keeps failing. Another conversation takes two minutes. Which
deserves attention first? This example takes you from an application snapshot
to the supporting events, then shows what the capture cannot tell you.

Everything runs locally without credentials or provider calls. The four
conversations are authored examples, not production records. They include
repeated message histories, overlapping tools, a delayed follow-up, and a
conversation with content capture disabled and one missing end timestamp.

## Load the snapshot

Install the package and the small JSON reader once:

```{r, eval=FALSE}
pak::pak(c("JamesHWade/scans", "jsonlite"))
```

```{r}
library(scans)
path <- system.file("extdata", "support-investigation.json", package = "scans")
bundle <- as_trajectory_otel(jsonlite::read_json(path))
trajectory_info(bundle)[, c("trajectory_id", "model")]
```

The fixture's source is `data-raw/investigation-fixture.R` in the repository.
The installed JSON is sufficient to repeat every calculation below.

## Find a pattern worth investigating

```{r}
findings <- scan_trajectories(bundle)
findings[, c("trajectory_id", "scan", "label")]
```

Only `otel/retry` is flagged. It has three tool errors, a repeated request,
and a suspicious loop. Those five findings describe **one affected
conversation**, not five failing conversations. They identify behavior to
review; they do not establish that the final answer was wrong.

Inspect the events that support the loop finding:

```{r}
loop <- findings[findings$scan == "suspicious_tool_loop", ]
events <- trajectory_events(bundle)
evidence <- events[events$event_id %in% loop$event_ids[[1]], ]
evidence[, c("event_id", "event_type", "name", "duration", "error")]
```

All three calls ask `lookup_order` for the same order. Each ends with
`order_service_timeout`. The transcript's final answer acknowledges that it
could not retrieve the shipment. A useful engineering question is whether
the workflow should stop repeating that request or use a recovery route.
This record alone does not tell us whether a retry policy was intentional,
or whether the order service was failing for other users.

## Separate waiting from recorded work

```{r}
measures <- measure_trajectories(bundle)
timing <- measures[measures$measure %in% c(
  "elapsed", "model_duration", "tool_duration",
  "recorded_work", "unattributed_elapsed"
), ]
timing[, c("trajectory_id", "measure", "value", "n_recorded", "n_total", "coverage")]
```

The follow-up conversation spans **126 seconds**, with **6 seconds** of
recorded model work and **120 seconds** outside recorded work. Its elapsed
time does not establish a slow model. The parallel example records **4
seconds** in model calls and **10 seconds** in tools, but **10 seconds** of
work after overlapping intervals are counted once. Adding its inclusive
durations would wrongly claim 14 seconds of wall time.

The limited capture has one valid four-second interval and one span without
an end timestamp. Recorded work is therefore partial. Its elapsed time and
time outside recorded work are unavailable. No transcript findings on that
conversation do not establish a successful or well-behaved interaction.

## Check the recording denominator

```{r}
usage <- measures[measures$measure %in% c("input_tokens", "output_tokens"), ]
usage[, c("trajectory_id", "measure", "value", "n_recorded", "n_total", "coverage")]
```

The retry conversation consumed 1,000 recorded input tokens across four
model calls. Its reconstructed final assistant turn carries only the latest
400 input tokens. `summarize_trajectories()` keeps its original turn-based
usage sums; use `measure_trajectories()` for the source-aware resource view
shared with the app. Cached input is part of input usage, not an additional
token count to add to it.

`complete` means every eligible **captured** record supplied a valid value.
It cannot prove that a source recorded every call, retained every trace, or
loaded the whole time window. The limited capture supplies input usage for
one of two calls and no output usage. Unknown usage is never converted to
zero.

```{r}
measures[measures$trajectory_id == "otel/retry" &
  measures$measure %in% c("first_chunk_latency", "first_token_latency", "cost"),
  c("measure", "value", "coverage", "limitation")]
```

This fixture has no first-chunk measurement. Where an OTel producer records
`gen_ai.response.time_to_first_chunk`, scans reports its median in seconds.
A chunk need not contain a token, so it is not labelled first-token latency.
Comparable cost is unavailable without currency, a price basis, and known
inclusion of child work. The old raw cost column remains available for
source-specific analysis.

## Open the same investigation in the app

Install the optional UI dependencies once, then open the same bundle:

```{r, eval=FALSE}
pak::pak(c("shiny", "bslib", "htmltools", "commonmark", "xml2"))
scans_app(list("Support assistant example" = bundle))
```

The application overview uses these measurements. Select **suspicious tool
loop**, then open `otel/retry`. The transcript and findings retain the event
identifiers shown above. Expand **Resource measurements** to inspect each
quantity, recording denominator, source, and limitation. Clear the pattern
filter to inspect the other three conversations. Filter changes reuse the
loaded snapshot.

The overview's recorded-work medians exclude trajectories with partial timing;
individual resource details still show their known intervals. Elapsed medians
exclude unavailable bounds. Recorded token medians can include partial sums,
so inspect coverage before interpreting a difference as an improvement.

## Bring your own completed evidence

| Source | Public entry point | Resource and capture boundary |
| --- | --- | --- |
| ellmer chat or turns (>= 0.4.2) | `as_trajectory_ellmer()` | Canonical turn usage and duration where supplied. A turn duration is not asserted to be one model call. |
| OTel GenAI / Connect | `as_trajectory_otel()` / `read_connect_traces()` | Call-level usage and intervals when recorded. Message capture and conversation IDs determine whether a meaningful transcript can be reconstructed. Keep read limits with the snapshot. |
| vitals evaluation (>= 0.3.0) | `as_trajectory_vitals()` | Outcome evaluations joined to solver trajectories; resource coverage follows the supplied turns. Reading development vitals logs requires its supported development API. |
| Deputy and dsprrr public results | `as_trajectory_deputy()` / `as_trajectory_dsprrr()` | Source execution and trace measurements remain distinct from model-call timing. dsprrr millisecond latency is normalized to canonical seconds. |
| Tempest public review | `as_trajectory_tempest()` | Product review evidence is not a full call trace. The bundled public review fixture has no canonical token counts or model durations. |
| Commons conversations | `as_trajectory_commons()` | Already-read conversation evidence; availability depends on what the upstream reader retained. |

Optional integration versions are specified in `DESCRIPTION` and exercised
through public producer objects in the test suite. This is a capture table,
not a claim that every release or every deployment records every field.
Continue with [plain ellmer and adapters](adapters.html),
[Connect capture](connect.html), or the [core scans tour](scans.html).

Reviewed, immutable example export is the next workflow stage tracked in
[#41](https://github.com/JamesHWade/scans/issues/41). This recipe stops at
inspection; it does not silently turn captured conversations into evaluation
data. The full community onboarding issue
[#45](https://github.com/JamesHWade/scans/issues/45) stays open for that stage.
