---
title: "Get started with scans"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with scans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("ellmer", quietly = TRUE)
)
```

scans works on the record an agent leaves behind. It never calls a model or
runs a tool. This article introduces the trajectory model, the tour from a
completed run to a finding, and what each built-in scan means.

## The trajectory model

A **trajectory** is one agent run: a chat, a Deputy result, a Connect
conversation. A **bundle** holds one or more trajectories and is the unit every
scans function accepts.

Inside a trajectory:

- **Turns** are the conversational units, each with a role such as `user`,
  `assistant`, or `system`.
- **Events** are what happened within turns: text content, tool calls, tool
  results, thinking, errors, and source-specific records. Every event has an
  identifier, a position in canonical order, and optional timing, usage, and
  status.
- **Evaluations** are outcome judgements joined to a trajectory, for example a
  vitals score.
- **Losses** record what an adapter could not carry over from its source. A
  loss is explicit so a reviewer knows what is missing rather than guessing.

A **finding** is the output of a scan: a deterministic statement about a
trajectory that names the events supporting it.

## From a completed run to a finding

```{r}
library(scans)

first_request <- ellmer::ContentToolRequest(
  "call-weather-1",
  "weather",
  list(city = "Detroit")
)
second_request <- ellmer::ContentToolRequest(
  "call-weather-2",
  "weather",
  list(city = "Detroit")
)

turns <- list(
  ellmer::UserTurn("What's the weather in Detroit?"),
  ellmer::AssistantTurn(list(first_request)),
  ellmer::UserTurn(list(ellmer::ContentToolResult(
    "Cloudy",
    request = first_request
  ))),
  ellmer::AssistantTurn(list(second_request))
)

bundle <- as_trajectory(turns)
bundle
```

`as_trajectory()` dispatches on the input's class. An ellmer `Chat` works the
same way, as do the classed results of the other supported frameworks.

### Look at the tables

Every accessor returns an ordinary tibble:

```{r}
trajectory_info(bundle)
trajectory_turns(bundle)[, c("turn_id", "role", "turn_index")]
trajectory_events(bundle)[, c("event_id", "event_type", "name", "status")]
```

`filter_trajectory_events()` selects events by trajectory, role, type, tool,
status, or metadata and adds the containing turn's context in dot-prefixed
columns:

```{r}
filter_trajectory_events(bundle, event_type = "tool_call")[
  ,
  c("event_id", "name", ".turn_role")
]
```

### Summarize

`summarize_trajectories()` returns one row per trajectory with counts of turns,
events, tool activity, unresolved calls, failures, token usage, and duration,
so a fleet of runs can be sorted and filtered before anyone opens a transcript:

```{r}
summarize_trajectories(bundle)[
  ,
  c("trajectory_id", "n_tool_calls", "n_unresolved_tool_calls")
]
```

### Scan

`scan_trajectories()` runs the built-in detectors and returns one finding per
problem:

```{r}
findings <- scan_trajectories(bundle)
findings[, c("severity", "label", "explanation")]
findings$event_ids
```

Each finding carries `trajectory_id`. Event-backed findings also carry
`event_id` and `event_ids`, so those findings join directly back to events.
Record-level findings instead identify the failed trajectory or turn through
`trajectory_id` or `turn_id`. Finding identifiers are stable across runs of
the same bundle.

## What the built-in scans mean

```{r}
scan_registry()
```

| Scan | It fires when |
|---|---|
| `unresolved_tool_call` | A tool was called and no result came back. |
| `unmatched_tool_result` | A result arrived with no preceding call to match it. |
| `ambiguous_tool_correlation` | A call identifier matches more than one call or result, so the pairing cannot be trusted. |
| `repeated_tool_call` | The same tool was called with the same arguments at least `repeat_threshold` times (default 2). Argument order does not matter. |
| `suspicious_tool_loop` | At least `loop_threshold` consecutive identical calls (default 3) suggest the agent is stuck. |
| `event_error` | An event recorded a failure. |
| `error_chain` | A failure followed an earlier one in the same trajectory, which often means the agent did not recover. |
| `turn_error` | A turn has a failed status, error text, or a truncating finish reason such as a token limit. |
| `trajectory_error` | The run itself ended with a failed status or an error, even when no event recorded it. |

Run a subset with `scans =`, and adjust the thresholds for agents whose normal
behaviour includes retries:

```{r}
scan_trajectories(
  bundle,
  scans = c("repeated_tool_call", "event_error"),
  repeat_threshold = 3L
)
```

## Open the review app

`scans_app()` renders the same bundle as a read-only Shiny app: a trajectory
list you can search, filter, and sort; a transcript rendered as chat with tool
activity collapsed into rows; and an evidence rail that links each finding to
the events behind it.

```{r, eval = FALSE}
scans_app(bundle)
```

Pass a named list to review more than one application. Entries may be bundles
or zero-argument functions that load one when the application is first
selected:

```{r, eval = FALSE}
scans_app(list(
  "Support assistant" = support_bundle,
  "Research assistant" = function() load_research_bundle()
))
```

To let reviewers record labels and notes, pass a `scans_annotations()` store.
Records append to a JSON-lines log and are never rewritten, so two reviewers
cannot overwrite each other and a revised judgement keeps the one before it:

```{r, eval = FALSE}
store <- scans_annotations(
  "reviews/annotations.jsonl",
  labels = c("looks right", "wrong answer", "tool failure", "follow up")
)
scans_app(bundle, annotations = store)

store$read()
```

## Where to next

- `vignette("connect")` covers reading traces from Posit Connect and deploying
  the review app there.
- `vignette("adapters")` describes what each adapter preserves and how to
  build a bundle from another framework.
